---
title: "hw03"
author: Mielle
output: github_document

---

 > Pick at least three of the tasks below (in the “Task menu” section) and attack each with a table and figure. For each table, make sure to include a relevant figure! Note that:

> dplyr should be your data manipulation tool
ggplot2 should be your visualization tool
Make observations about what your tables/figures show and about the process.

> Also useful for you to add to your “cheat sheet” are notes on difficulties/oddities. For example, which figures are easy/hard to make, which data formats make better inputs for plotting functions vs. for human-friendly tables.

> If relevant, give credit to your sources, whether it’s a blog post, a fellow student, an online tutorial, etc. This is also valuable “cheat sheet” info for future-you.


Relax about the following things:

Tidying/reshaping is NOT your assignment. Many of your tables will be awkwardly shaped in the report. That’s OK.
Table beauty is not a big deal. Simply printing to “screen” is fine. You could also try the knitr::kable() function. Assuming my_df is a data.frame, here’s an R chunk that should print it as a decent-looking table:
``{r results = 'asis'}
knitr::kable(my_df)
``
For all things, graphical and tabular, if you’re dissatisfied with a result, discuss the problem, what you’ve tried and move on.
Your figure does not have to depict every single number from the data aggregation result. Use your judgement. It just needs to complement the table, add context, and allow for some sanity checking both ways.

## MENU
- Get the maximum and minimum of GDP per capita for all continents.

- Look at the spread of GDP per capita within the continents.

- Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.

- How is life expectancy changing over time on different continents?

- Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

- Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.

## Getting started
```{r load packages}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(scales))

```



## How is life expectancy changing over time on different continents?


```{r lifeExp over time}
lifetime <- gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mnlife = mean(lifeExp), mdlife = median(lifeExp))

lifetime
```

Now, is it better to use the mean or median to represent this data? Let's check if the data is normally distributed. 

```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(binwidth = 1, fill = "light blue") +
  xlab("Count") +
  ylab("Life expectancy") +
  ggtitle("Frequency distribution of global life expectancy")

```

It's pretty clear our data isn't normally distributed, so it's more appropriate to use the median to represent life expectancy. 



```{r}
ggplot(lifetime, aes(year, mdlife, color = continent)) +
  geom_line() +
  scale_color_brewer(palette = "Dark2") +
  xlab("Year") +
  ylab("Life expectancy") +
  ggtitle("Worldwide life expectancy over time")
    
```


## Look at the min + max of GDP per capita within the continents

make bar graph with min + max + median GDP on each continent (figure out how to get into ggplot as can't pipe it!)

```{r continent summary stats - gdp}
spread <- gapminder %>% 
  group_by(continent) %>% 
  summarise(min = min(gdpPercap), md = median(gdpPercap), max = max(gdpPercap))

spread
 
```

```{r continent gdppercap boxplot}
gdp <- ggplot(gapminder, aes(continent, gdpPercap, fill = continent))

ggplot(gapminder, aes(continent, gdpPercap, fill = continent)) +
  geom_boxplot() +
  scale_y_log10() +
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("GDP per capita")
```
```{r}
#geom_col for stacked bar graph
```



## Find countries with interesting stories 

#### Correlation of GDP per capita and life expectancy

First, let's look at the relationship between GDP per capita and life expectancy, using a basic scatterplot. 
```{r scatterplot of gdpPercap + lifeExp}
ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent, alpha = 0.7)) +
  geom_point() +
  scale_x_log10() +
  scale_color_brewer(palette = "Dark2") +
  xlab("GDP per capita") +
  ylab("Life expectancy") +
  ggtitle("Life expectancy vs. GDP per capita")

```


From this scatter plot, we can see that typically, GDP per capita and life expectancy appear positively correlated (as one increases, the other increases). Is there a country where this does not hold true? Can we get GDP increase with life expectancy decrease? 

#### Simultaneous GDP per capita increase and life expectancy decrease

Calculate the rate of change in life expectancy and GDP per capita between observations for each country.

```{r Change in lifeExp and gdpPercap}
change <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdpchange = gdpPercap - lag(gdpPercap)) %>% 
  mutate(lifechange = lifeExp - lag(lifeExp)) %>% 
  transmute(gdpchange, lifechange, year)

change
```
 

Select observations where GDP remained the same or increased, and life expectancy decreased. 

```{r gdp up, lifeExp down}
change %>% 
  filter(gdpchange >= 0, lifechange < 0)
```

Out of the entire dataset, this has only occured 54 times! Our assumption that GDP per capita and life expectancy will increase together was largely correct, with a few exceptions. 



``{r}
# having trouble graphing because ggplot2 doesn't group the way that dplyr does
change %>% 
  filter(gdpchange >= 0, lifechange < 0) %>% 
  ggplot(aes(gdpchange, lifechange)) +
  geom_bar(aes(stat = identity, position = dodge))
```

We can tell that this phenomena is rare, but let's see how rare it actually is.
RELATIVE FREQUENCY? would need to filter out years that have NAs and then take the 54 instances out of that. 
```{r}

```

#### Simultaneous population increase and life expectancy decrease

Now, let's see if population growth and life expectancy sometimes change in opposite directions. 

```{r pop up, lifeExp down}
popup <-gapminder %>% 
  group_by(country) %>% 
  mutate(popchange = pop - lag(pop)) %>% 
  mutate(lifechange = lifeExp - lag(lifeExp)) %>% 
  filter(popchange >= 0, lifechange < 0) %>% 
  transmute(popchange, lifechange, year, continent)

popup
```

```{r}
ggplot(popup, aes(continent, fill = continent)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("Count") +
  ggtitle("Frequency of instances of increasing population and decreasing life expectancy")
```


Yes, population increases do happen concurrently with life expectancy decreases, and at almost twice the rate as the same phenomenon with GDP per capita and population. Perhaps increasing population means more pressure on resources, which decreases the health and therefor life expectancy of the population. This phenomenon is most common in countries from Africa. 




#### Simultaneous population decrease and life expectancy increase 

Now, let's check the opposite-- do life expectancy increases ever coincide with population decreases?

```{r pop down, lifeExp up}
popdown <- gapminder %>% 
  group_by(country) %>% 
  mutate(popchange = pop - lag(pop)) %>% 
  mutate(lifechange = lifeExp - lag(lifeExp)) %>% 
  filter(popchange < 0, lifechange >= 0) %>% 
  transmute(popchange, lifechange, year, continent)

popdown
```
MAYBE EXCLUDE SCATTER PLOT, OR NEED TO DISCUSS IT
```{r scatter plot of lifechange, popchange}
ggplot(popdown, aes(lifechange, popchange, color = continent)) +
  scale_color_brewer(palette = "Dark2") +
  geom_point()
```

Occasionally, yes! With only 36 observations, this is our rarest find yet. Let's explore if there's a spatial pattern here. 

```{r count countries by continent}
popdown %>% 
  group_by(continent) %>% 
  count(continent)

```

Let's do a quick visualization to compare. 
```{r popdown histogram by continent}
ggplot(popdown, aes(continent, fill = continent)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("Count") +
  ggtitle("Frequency of instances of increasing life expectancy and decreasing population")
```

Yes, even without mapping there is a clear spatial pattern-- the vast majority of observations where population decreases as life expectancy increases are in Europe. Ideally, we would look at the relative frequency of each continent in the dataset to see if there is truly a disproportionate number of observations. However, we know that Europe does not have significantly more countries than all the other continents, so even without a statistical comparison it's pretty clear there is a spatial dimension. 




Let's see if there's a temporal pattern as well. This isn't working though! 

```{r popdown histogram by year}
popdown %>% 
  count(year) %>% 
  arrange(year)

ggplot(popdown, aes(year)) +
  geom_bar() +
  xlab("Year") +
  ylab("Count") +
  ggtitle("Temporal pattern of increasing life expectancy and decreasing population ")
```


Yes, there is a clear temporal pattern to this phenomenon. It appears that over time, increasing life expectancy combined with decreasing population has become more frequent. 

