---
title: "Exploring gapminder data with dplyr and ggplot2"
author: Mielle Michaux
output:
  github_document:
    toc: true
    toc_depth: 4
---

## Brief intro 

I'm using dplyr and ggplot2 to manipulate and visualize the gapminder dataset, which includes population, life expectancy, and GDP per capita by country from 1952 to 2007. I've done a fairly brief exploration of minimum and maximum GDP and changing life expectancy over time, and a more in-depth look at some oddball cases that go against general trends in the data. 


## Getting started
```{r load packages}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(scales))

```


## 1. How is life expectancy changing over time on different continents?

#### Using the mean and median to compare life expectancy 

```{r lifeExp over time}
lifetime <- gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mnlife = mean(lifeExp), mdlife = median(lifeExp))

lifetime
```

Now, is it better to use the mean or median to represent this data? Let's check if the data is normally distributed. 

```{r histogram of life expectancy}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(binwidth = 1, fill = "dark blue") +
  xlab("Life expectancy") +
  ylab("Count") +
  ggtitle("Frequency distribution of global life expectancy")

```

It's pretty clear our data isn't normally distributed, so it's more appropriate to use the median to represent life expectancy. 

#### Plotting global trends in median life expectancy over time

First, we can plot median life expectancy for 1952 and 2007, and compare those. 

```{r 1952 median life expectancy (bar graph)}
lifetime %>% 
  filter(year == 1952) %>% 
  ggplot(aes(continent, mdlife, fill = continent)) +
  geom_col() +  
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("Median life expectancy") +
  ggtitle("Worldwide life expectancy, 1952")
```

In 1952, median life expectancy for Oceania and Europe was fairly similar and at the high end, Africa and Asia similar and at the low end, and the Americas had a median somewhere in between. 

```{r 2007 median life expectancy (bar graph)}
lifetime %>% 
  filter(year == 2007) %>% 
  ggplot(aes(continent, mdlife, fill = continent)) +
  geom_col() +  
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("Median life expectancy") +
  ggtitle("Worldwide life expectancy, 2007")
```

Looks like in 2007 Oceania, Europe, Asia, and the Americas are fairly close for median life expectancy with Africa close to 20 years lower than the rest. 


Now, let's look at median life expectancy for every year the data was collected. 

The graph below shows median life expectancy over time for each continent. Because ggplot2 doesn't group variables the way that dplyr does, you'll notice that I wasn't always able to directly pipe my tables created with dplyr into ggplot for visualization-- sometimes, I had to go about it a different way in order to graph the same thing. I would have liked to add another line with global median life expectancy, but couldn't get it to work. 


```{r md lifeExp over time}
ggplot(lifetime, aes(year, mdlife, color = continent)) +
  geom_line() +
  scale_color_brewer(palette = "Dark2") +
  xlab("Year") +
  ylab("Median life expectancy") +
  ggtitle("Worldwide life expectancy, 1952 - 2007")
    
```

Looks like all continents generally have a trend of increasing life expectancy over time, with Europe and Oceania following a similar trend and maintaining the highest median life expectancy. Asia and the Americas have a more dramatic increase over time, starting much lower than Europe and Oceania but closing the gap to within about five years by 2007. Median life expectancy in Africa is the lowest throughout, with a less dramatic gain over time-- it appears that the gap between Africa and the rest of the continents has widened over time. 


#### Plotting global change in life expectancy
  
Another way to look at change in life expectancy is to calculate the difference in life expectancy between 1952 and 2007 for each continent. 

You'll notice in the table below that I've gone about this in a fairly messy way. Rather than subtracting the life expectancy 1952 from the life expectancy in 2007, I took advantage of the fact that it increased for every continent and simply filtered for the desired years and then subtract the maximum median life expectancy from the minimum by continent. Quick and dirty, but not ideal for replicating this if the data were different. 

```{r calculate increase in life expectancy }
lifeincr <- gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mdlife = median(lifeExp)) %>% 
  filter(year == 2007 | year == 1952) %>% 
  mutate(change = max(mdlife) - min(mdlife)) %>% 
  filter(year == 2007)
  
lifeincr

```

Here is the table above, represented as a bar graph. Asia had the largest increase in median life expectancy by far, with an almost 30-year gain. Oceania and Europe made the smallest gains, but as we already know that those continents have maintained the highest median life expectancy, their lack of gain isn't necessarily an indicator of anything negative. This is a great example of why it's important to examine data in more than one way-- if we only had this bar graph to analyze, our takeaway for Oceania and Europe would probably be pretty different. 

```{r}
ggplot(lifeincr, aes(continent, change, fill = continent)) +
  geom_col() +
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("Change in life expectancy") +
  ggtitle("Worldwide change in life expectancy, 1952 - 2007")
```



## 2: Look at the minimum and maximum of GDP per capita within the continents

make bar graph with min + max + median GDP on each continent (figure out how to get into ggplot as can't pipe it!)

#### Minimum and maximum GDP per capita with a boxplot 
We can use a boxplot to examine the minimum, maximum, and median values of GP per capita for each continent. 

```{r continent gdppercap boxplot}
gdp <- ggplot(gapminder, aes(continent, gdpPercap, fill = continent))

gdp +
  geom_boxplot() +
  scale_y_log10() +
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("GDP per capita")
```



```{r}
minmax <- gapminder %>% 
  group_by(continent) %>% 
  summarise(min = min(gdpPercap), md = median(gdpPercap), max = max(gdpPercap)) 

minmax

```

Let's see a graph of the minimum GDP per capita value for each continent. 

```{r}
ggplot(minmax, aes(continent, min, fill = continent)) +
  geom_col() +
  scale_fill_brewer(palette = "Dark2") +
  ggtitle("Minimum GDP per capita by continent, 1952-2007") +
  xlab("Continent") +
  ylab("GDP per capita value")
```

Now, a graph of the maximum GDP per capita value for each continent. 

```{r}
ggplot(minmax, aes(continent, max, fill = continent)) +
  geom_col() +  
  ggtitle("Maximum GDP per capita by continent, 1952-2007") +
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("GDP per capita value")
```

Because a common y-scale would have obscured most of the differences between continents for minimum GDP per capita, I refrained from faceting the minimum and maximum GDP graphs to display side by side. I generally avoid faceting graphs with different y scales, as the facet layout can give the impression that they are on the same scale if the audience isn't paying close attention. 

## 3: Find countries with interesting stories 

#### Correlation of GDP per capita and life expectancy

First, let's look at the relationship between GDP per capita and life expectancy, using a basic scatterplot. 

```{r scatterplot of gdpPercap + lifeExp}
ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent, alpha = 0.7)) +
  geom_point() +
  scale_x_log10() +
  scale_color_brewer(palette = "Dark2") +
  xlab("GDP per capita") +
  ylab("Life expectancy") +
  ggtitle("Life expectancy vs. GDP per capita")

```

From this scatter plot, we can see that typically, GDP per capita and life expectancy appear positively correlated (as one increases, the other increases). Is there a country where this does not hold true? Can we get GDP increase with life expectancy decrease? 

#### Simultaneous GDP per capita increase and life expectancy decrease

To look at GDP and life expectancy changes, we need to calculate the rate of change in life expectancy and GDP per capita between observations for each country.

```{r Change in lifeExp and gdpPercap}
change <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdpchange = gdpPercap - lag(gdpPercap)) %>% 
  mutate(lifechange = lifeExp - lag(lifeExp)) %>% 
  transmute(gdpchange, lifechange, year, continent)

change
```

Next, select observations where GDP remained the same or increased, and life expectancy decreased. 

```{r gdp up, lifeExp down}
diffchange <- change %>% 
  filter(gdpchange >= 0, lifechange < 0)

diffchange
```

Out of the entire dataset, this has only occurred 54 times! Our assumption that GDP per capita and life expectancy will increase together was largely correct, with these 54 exceptions. 

```{r histogram of gdp up, lifeexp down by continent}
ggplot(diffchange, aes(continent, fill = continent)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") + 
  ggtitle("Frequency of increasing GDP per capita and decreasing life expectancy") +
  xlab("Continent") +
  ylab("Count")
```

Most of the cases of increasing GDP per capita and decreasing life expectancy occurred in Africa, with a significant subset in Europe. 


```{r histogram of gdp up, lifeexp down by year}
ggplot(diffchange, aes(year)) +
  geom_histogram(binwidth = 5, fill = "dark blue") +
  ggtitle("Frequency of increasing GDP per capita and decreasing life expectancy") +
  xlab("Year") +
  ylab("Count")

```

Over time, GDP per capita increase and life expectancy decrease became more common, and then dropped off sharply with the 2007 dataset. 

#### Simultaneous population increase and life expectancy decrease

Now, let's see if population growth and life expectancy sometimes change in opposite directions. 

```{r pop up, lifeExp down}
popup <-gapminder %>% 
  group_by(country) %>% 
  mutate(popchange = pop - lag(pop)) %>% 
  mutate(lifechange = lifeExp - lag(lifeExp)) %>% 
  filter(popchange >= 0, lifechange < 0) %>% 
  transmute(popchange, lifechange, year, continent)

popup
```

```{r bar graph of incr pop, decr lifeExp}
ggplot(popup, aes(continent, fill = continent)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("Count") +
  ggtitle("Frequency of instances of increasing population and decreasing life expectancy")
```

Yes, population increases do happen concurrently with life expectancy decreases, and at almost twice the rate as the same phenomenon with GDP per capita and population. Perhaps increasing population means more pressure on resources, which decreases the health and therefor life expectancy of the population. This phenomenon is most common in countries from Africa. 


#### Simultaneous population decrease and life expectancy increase 

Now, let's check the opposite-- do life expectancy increases ever coincide with population decreases?

```{r pop down, lifeExp up}
popdown <- gapminder %>% 
  group_by(country) %>% 
  mutate(popchange = pop - lag(pop)) %>% 
  mutate(lifechange = lifeExp - lag(lifeExp)) %>% 
  filter(popchange < 0, lifechange >= 0) %>% 
  transmute(popchange, lifechange, year, continent)

popdown
```

Occasionally, yes! With only 36 observations, this is our rarest find yet.

Let's check out a scatter plot to see how our observations vary in population and life expectancy change. Here, I've labelled the observations that are visually quite different than the rest of the group. 

```{r scatter plot of lifechange, popchange}
ggplot(popdown, aes(lifechange, popchange, color = continent, label = country)) +
  scale_color_brewer(palette = "Dark2") +
  geom_point() +
  xlab("Change in life expectancy") +
  ylab("Change in population") +
  geom_text(aes(label = ifelse (lifechange > 3 | popchange < -1500000, as.character(country), '')), vjust = -1, hjust = .7) +
  geom_text(aes(label = ifelse (lifechange > 3 | popchange < -1500000, as.character(year), '')), vjust = .5, hjust = 1.2)
  
```

Looks like most decreases in population in this data set are under 500,000, and most changes in life expectancy are an increase of less than 2 1/2 years. 



 Now, let's explore if there's a spatial pattern. 

```{r count countries by continent}
popdown %>% 
  group_by(continent) %>% 
  count(continent)

```

Let's do a quick visualization to compare. 
```{r popdown histogram by continent}
ggplot(popdown, aes(continent, fill = continent)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  xlab("Continent") +
  ylab("Count") +
  ggtitle("Frequency of instances of increasing life expectancy and decreasing population")
```

Yes, even without mapping there is a clear spatial pattern-- the vast majority of observations where population decreases as life expectancy increases are in Europe. Ideally, we would look at the relative frequency of each continent in the dataset to see if there is truly a disproportionate number of observations. However, we know that Europe does not have significantly more countries than all the other continents, so even without a statistical comparison it's pretty clear there is a spatial dimension. 


Let's see if there's a temporal pattern as well. 

```{r popdown histogram by year}
popdown %>% 
  count(year) %>% 
  arrange(year)

ggplot(popdown, aes(year)) +
  geom_bar(fill = "dark blue") +
  xlab("Year") +
  ylab("Count") +
  ggtitle("Temporal pattern of increasing life expectancy and decreasing population ")
```

Yes, there is a clear temporal pattern to this phenomenon. It appears that over time, increasing life expectancy combined with decreasing population has become more frequent.


## Special thanks to: 

- [Bar graph help](https://ggplot2.tidyverse.org/reference/geom_bar.html)

- [Labelling points in a scatterplot](https://stackoverflow.com/questions/15624656/label-points-in-geom-point)

- [Github-flavored markdown refresher #1](https://guides.github.com/features/mastering-markdown/)

- [Github-flavored markdown refresher #2](https://rmarkdown.rstudio.com/github_document_format.html#table_of_contents)

