---
title: "hw03"
author: Mielle
output: github_document
---

 > Pick at least three of the tasks below (in the “Task menu” section) and attack each with a table and figure. For each table, make sure to include a relevant figure! Note that:

> dplyr should be your data manipulation tool
ggplot2 should be your visualization tool
Make observations about what your tables/figures show and about the process.

> Also useful for you to add to your “cheat sheet” are notes on difficulties/oddities. For example, which figures are easy/hard to make, which data formats make better inputs for plotting functions vs. for human-friendly tables.

> If relevant, give credit to your sources, whether it’s a blog post, a fellow student, an online tutorial, etc. This is also valuable “cheat sheet” info for future-you.


Relax about the following things:

Tidying/reshaping is NOT your assignment. Many of your tables will be awkwardly shaped in the report. That’s OK.
Table beauty is not a big deal. Simply printing to “screen” is fine. You could also try the knitr::kable() function. Assuming my_df is a data.frame, here’s an R chunk that should print it as a decent-looking table:
```{r results = 'asis'}
knitr::kable(my_df)
```
For all things, graphical and tabular, if you’re dissatisfied with a result, discuss the problem, what you’ve tried and move on.
Your figure does not have to depict every single number from the data aggregation result. Use your judgement. It just needs to complement the table, add context, and allow for some sanity checking both ways.

## MENU
- Get the maximum and minimum of GDP per capita for all continents.

- Look at the spread of GDP per capita within the continents.

- Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.

- How is life expectancy changing over time on different continents?

- Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

- Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.


```{r load packages}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```


# Look at the spread of GDP per capita within the continents.
make bar graph with min + max + median GDP on each continent (figure out how to get into ggplot as can't pipe it!)

```{r continent summary stats - gdp}
gapminder %>% 
  group_by(continent) %>% 
  summarise(min = min(gdpPercap), md = median(gdpPercap), max = max(gdpPercap))
 
```

```{r}
ggplot()
```



# Find countries with interesting stores 

Typically, GDP and life expectancy increase together. Is there a country where this does not hold true? Can we get GDP increase with life expectancy decrease? 

rate of change for gdp 
rate of change for lifeexp 

Calculate the rate of change in life expectancy and GDP per capit between observations for each country.

```{r Change in lifeExp and gdpPercap}
change <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdpchange = gdpPercap - lag(gdpPercap)) %>% 
  mutate(lifechange = lifeExp - lag(lifeExp)) %>% 
  transmute(gdpchange, lifechange, year)

change
```
 

Select observations where GDP remained the same or increased, and life expectancy decreased. 

```{r gdp up, lifeExp down}
change %>% 
  filter(gdpchange >= 0, lifechange < 0)

```

Out of the entire dataset, this has only occured 54 times! Our assumption that GDP per capita and life expectancy will increase together was largely correct, with a few exceptions. 



```{r}
# having trouble graphing because ggplot2 doesn't group the way that dplyr does
change %>% 
  filter(gdpchange >= 0, lifechange < 0) %>% 
  ggplot(aes(gdpchange)) +
  geom_bar()
```

We can tell that this phenomena is rare, but let's see how rare it actually is.

```{r}

```


Now, let's see if population change and life expectancy sometimes change in opposite directions. 

```{r pop up, lifeExp down}
gapminder %>% 
  group_by(country) %>% 
  mutate(popchange = pop - lag(pop)) %>% 
  mutate(lifechange = lifeExp - lag(lifeExp)) %>% 
  filter(popchange >= 0, lifechange < 0) %>% 
  transmute(popchange, lifechange, year)
```

Yes, population increases do happen concurrently with life expectancy decreases, and at almost twice the rate as the same phenomenon with GDP per capita and population. Perhaps increasing population means more pressure on resources, which decreases the health and therefor life expectancy of the population. 

Now, let's check the opposite-- do life expectancy increases ever coincide with population decreases?

```{r pop down, lifeExp up}
popdown <- gapminder %>% 
  group_by(country) %>% 
  mutate(popchange = pop - lag(pop)) %>% 
  mutate(lifechange = lifeExp - lag(lifeExp)) %>% 
  filter(popchange < 0, lifechange >= 0) %>% 
  transmute(popchange, lifechange, year, continent)

popdown
```

Occasionally, yes! With only 36 observations, this is our rarest find yet. Let's explore if there's a spatial pattern here. 

```{r count countries by continent}
popdown %>% 
  group_by(continent) %>% 
  count(continent)
```

Yes, even without mapping there is a clear spatial pattern-- the vast majority of observations where population decreases as life expectancy increases are in Europe. Ideally, we would look at the relative frequency of each continent in the dataset to see if there is truly a disproportionate number of observations. However, we know that Europe does not have significantly more countries than all the other continents, so even without a statistical comparison it's pretty clear there is a spatial dimension. 




Let's see if there's a temporal pattern as well. THis isn't working though! 

```{r count countries by year}
popdown %>% 
  count(year) %>% 
  arrange(year)
```




